Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-12-12 scagnostics.Rmd
Updated spiral guided tour example text accordingly.

The spiral is the most challenging structure to detect amongst the distribution considered here (compare also Posses results?). Here we explore the role of dimensionality in the index optimization, using the skinny index that was found to be sensitive to this structure.
We use a setup similar to that introduced for detecting the pipe structure, i.e.\ we use the search\_better optimization of the guided tour, disabling the cooling and with a large search window, while sampling a large number of random projections (up to 5000) for each step.
We consider distributions in d=4,6,8 dimensions and compare features found in the tour.
In d=4 dimensions the optimization identifies a view showing the spiral after 33 steps, and we limit the optimizer to consider only 33 steps for the study in 6 and 8 dimensions for direct comparison. Note however that the spiral is typically also not identified with much larger number of steps, the limiting factor being rather the number of random projections sampled and rotation non-invariance of the index prohibiting the convergence, in particular in higher dimensions.
Results are shown in the top panel Figure \ref{fig:findspiral}, showing the pairwise distance of the 33 planes found in 4, 6 and 8 dimensions (left), and comparing the evolution of the distance from the plane containing the spiral view with progression of the optimization (right). We see that the pairwise distances grow with d (what do we learn from that?). Moreover we find that the distance to the special view is far from monotonically decreasing towards zero, and large jumps are possible. Note that the spiral itself is approximately rotation invariant, but partial views are not. At medium distances from the special view we often observe a jump to higher distances as consequence of in-plane rotation dependence for such partial views. This stabilizes closer to the special view, however it relies on randomly identifying a projection within this angle. As a consequence increasing dimensionality dramatically reduces the chance of identifying structures with high index values.

```{r findspiral, out.width="\\textwidth", fig.cap="Guided tour optimising the skinny index for the Sprial dataset with 1000 datapoints, with d = 4, 6, 8. The top left plot shows the distribution of pairwise distances between planes obtained via the guided tour, the top right shows the evolution of distance to the plane containing the spiral as the index is being optimised. The bottom row shows the final view obtained for d = 4 (left), d = 6 (middle) and d = 8 (right) after 33 steps in the guided tour.", results="hide"}

>>>>

The spiral is the most challenging structure to detect amongst the distribution considered here (compare also Posses results?). Here we explore the role of dimensionality in the index optimization, using the skinny index that was found to be sensitive to this structure.
We use a setup similar to that introduced for detecting the pipe structure, i.e.\ we use the search\_better optimization of the guided tour, disabling the cooling and with a large search window, while sampling a large number of random projections (up to 5000) for each step.
We consider distributions in d=4,5,6 dimensions (subsets of the spiral dataset) and compare features found in the tour.
In d=4 dimensions the optimization identifies a view showing the spiral after 39 steps. For larger values of d the optimization does not uncover the spiral view, and we limit the optimizer to 100 steps to compare the so obtained bases to the results found in four dimensions.
Results are shown in the top panel Figure \ref{fig:findspiral}. The top left is showing the pairwise distances between all planes found by the optimization, note that while d=5,6 contains 100 planes, d=4 only contains 39 planes. The top right is then comparing the evolution of the distance from the plane containing the spiral view with progression of the optimization. We see that the pairwise distances grow with d (what do we learn from that?). Moreover we find that the distance to the special view is far from monotonically decreasing towards zero, and large jumps are possible as a consequence of rotation dependence. Note that the spiral itself is approximately rotation invariant, but partial views are not. At medium distances from the special view we often observe a jump to higher distances as consequence of in-plane rotation dependence for such partial views. This stabilizes closer to the special view, however it relies on randomly identifying a projection within this angle. As a consequence increasing dimensionality dramatically reduces the chance of identifying structures with high index values.

```{r findspiral, out.width="\\textwidth", fig.cap="Guided tour optimising the skinny index for the Sprial dataset with 1000 datapoints, with d = 4, 5, 6. The top left plot shows the distribution of pairwise distances between planes obtained via the guided tour, the top right shows the evolution of distance to the plane containing the spiral as the index is being optimised. The bottom row shows the final view obtained for d = 4 (left), d = 5 (middle) and d = 6 (right).", results="hide"}
